Topic models

### 2021-03-01 topic_models.Rmd
Submitted revision

```{r setup, include=FALSE}
library(yaml)
knitr::opts_chunk$set(echo = FALSE)
lda_dims <- read_yaml("Study1_lda_dims.txt")
dtm_top <- read_yaml("study1_dtm_top.yml")
df <- readRDS("study1_df.RData")
number_docs_words <- yaml::read_yaml("study1_number_docs_words.txt")
```

@@

```{r, warning=FALSE, message=FALSE}
lda_dims_2 <- read_yaml("Study2_lda_dims.txt")
```
The corpus for this second analysis consisted of the abstracts of the selected articles.
To perform feature extraction,
we first applied the natural language processing technique "part-of-speech tagging" (POS-tagging), which identifies a word's grammatical function within the sentence context.
Because our analysis sought to identify phenomena,
we retained only nouns (to capture terms like "emotion") and adjectives (to capture the "mental" in "mental health").
Retaining nouns and adjectives generally helps derive more interpretable text mining models [@martinMoreEfficientTopic2015].
Finally, we used stemming to reduce the retained terms to their root form.
The resulting corpus consisted of 15587 terms in 4414 documents.

>>>>

```{r setup, include=FALSE}
library(yaml)
library(stringr)
library(udpipe)
library(igraph)
library(wordcloud)
library(Matrix)
library(yaml)
library(Rmpfr)
library(topicmodels)
library(udpipe)
library(slam)
library(tidytext)
library(ggplot2)
library(textrank)
source("word_functions.R")
knitr::opts_chunk$set(echo = FALSE)

# LDA analysis ------------------------------------------------------------

# Functions:
harmonicMean <- function(logLikelihoods, precision = 2000L) {
  llMed <- median(logLikelihoods)
  as.double(llMed - log(mean(exp(-mpfr(logLikelihoods,
                                       prec = precision) + llMed))))
}
BIC <- function(ll, p, n){
  -2 * ll + p * log(n)
}
entropy <- function(post_prob){
  1 + (1/(nrow(post_prob) * 
            log(ncol(post_prob)))) * (sum(rowSums(post_prob * 
                                                    log(post_prob + 1e-12))))
}

# Preprocessing -----------------------------------------------------------
df <- readRDS("study1_df.RData")

# Frequency of word by doc
nounbydoc <- df[, list(freq = .N), by = list(doc_id = doc, term = word_coded)]
# Set frequency to 1; we're not interpreting word frequency, only occurrence
nounbydoc$freq <- 1

df_lda <- nounbydoc #[nounbydoc$term %in% names(dict), ]
df_lda <- df_lda %>%
  bind_tf_idf(term, doc_id, freq)
summary(df_lda$tf_idf)
df_lda <- df_lda[order(df_lda$tf_idf),]

select_these <- df_lda[!duplicated(df_lda$term), ]
select_these <- select_these$term[select_these$tf_idf >= median(select_these$tf_idf)]
df_lda <- df_lda[df_lda$term %in% select_these, ]

dtm <- udpipe::document_term_matrix(document_term_frequencies(df_lda))

# dtm <- as.simple_triplet_matrix(dtm)
# dim(dtm)
# summary(col_sums(dtm))
# term_tfidf <- tapply(dtm$v/row_sums(dtm)[dtm$i], dtm$j, mean) * log2(dim(dtm)[1]/col_sums(dtm > 0))
# summary(term_tfidf)

yaml::write_yaml(dim(dtm), file = "Study1_lda_dims.txt")

# Build topic models

seqk <- seq(2, 20, 1)
burnin <- 1000
iter <- 1000
keep <- 50
set.seed(44773)
res_lda <- lapply(seqk, function(k) {
  topicmodels::LDA(
    dtm,
    k = k,
    method = "Gibbs",
    control = list(
      burnin = burnin,
      iter = iter,
      keep = keep
    )
  )
})

ll <- sapply(res_lda, function(x){harmonicMean(x@logLiks[-c(1:(burnin/keep))])})

K = seqk
N = nrow(dtm)
M = ncol(dtm)

parameters <- K*(M-1)+N*(K-1)
N <- nrow(dtm)
bics <- BIC(ll, parameters, N)

entropies <- sapply(res_lda, function(x){entropy(x@gamma)})

p <- ggplot(data.frame(K = seqk, Entropy = entropies), aes(x = K, y = Entropy)) + geom_path() +
  xlab('Number of topics') +
  scale_y_continuous(limits = c(0,1)) +
  theme_bw()

ggsave("study1_entropies.png", p, device = "png")
ggsave("study1_entropies.svg", p, device = "svg")

p <- ggplot(data.frame(K = seqk, ll = ll), aes(x = K, y = ll)) + geom_path() +
  geom_vline(xintercept = (which.max(ll)+1), linetype = 2) +
  xlab('Number of topics') +
  geom_smooth(method = "lm", formula = y~log(x), se = FALSE)+
  theme_bw()

ggsave("study1_ll.png", p, device = "png")
ggsave("study1_ll.svg", p, device = "svg")

p <- ggplot(data.frame(K = seqk, BIC = bics), aes(x = K, y = BIC)) + geom_path() +
  geom_vline(xintercept = (which.min(bics)+1), linetype = 2) +
  xlab('Number of topics') +
  geom_smooth(method = "lm", formula = y~x, se = FALSE)+
  theme_bw()

ggsave("study1_BIC.png", p, device = "png")
ggsave("study1_BIC.svg", p, device = "svg")

lda_dims <- read_yaml("Study1_lda_dims.txt")
dtm_top <- read_yaml("study1_dtm_top.yml")
df <- readRDS("study1_df.RData")
number_docs_words <- yaml::read_yaml("study1_number_docs_words.txt")
```

@@

```{r, warning=FALSE, message=FALSE}
df <- readRDS("study2_df.RData")
# Functions:
harmonicMean <- function(logLikelihoods, precision = 2000L) {
  llMed <- median(logLikelihoods)
  as.double(llMed - log(mean(exp(-mpfr(logLikelihoods,
                                       prec = precision) + llMed))))
}
BIC <- function(ll, p, n){
  -2 * ll + p * log(n)
}
entropy <- function(post_prob){
  1 + (1/(nrow(post_prob) * 
            log(ncol(post_prob)))) * (sum(rowSums(post_prob * 
                                                    log(post_prob + 1e-12))))
}

# Preprocessing -----------------------------------------------------------

# Frequency of word by doc
nounbydoc <- df[df$upos %in% c("NOUN", "ADJ"), list(freq = .N), by = list(doc_id = doc_id, term = lemma)]
df_lda <- nounbydoc #[nounbydoc$term %in% names(dict), ]
df_lda <- df_lda %>%
  bind_tf_idf(term, doc_id, freq)
summary(df_lda$tf_idf)

select_these <- df_lda[!duplicated(df_lda$term), ]
select_these <- select_these$term[select_these$tf_idf >= median(select_these$tf_idf)]
df_lda <- df_lda[df_lda$term %in% select_these, ]

dtm <- udpipe::document_term_matrix(document_term_frequencies(df_lda))
yaml::write_yaml(dim(dtm), file = "Study2_lda_dims.txt")

# Build topic models
seqk <- seq(2, 20, 1)
burnin <- 1000
iter <- 1000
keep <- 50
set.seed(44773)
res_lda <- lapply(seqk, function(k) {
  topicmodels::LDA(
    dtm,
    k = k,
    method = "Gibbs",
    control = list(
      burnin = burnin,
      iter = iter,
      keep = keep
    )
  )
})

ll <- sapply(res_lda, function(x){harmonicMean(x@logLiks[-c(1:(burnin/keep))])})

K = seqk
N = nrow(dtm)
M = ncol(dtm)

parameters <- K*(M-1)+N*(K-1)
N <- nrow(dtm)
bics <- BIC(ll, parameters, N)

entropies <- sapply(res_lda, function(x){entropy(x@gamma)})

p <- ggplot(data.frame(K = seqk, Entropy = entropies), aes(x = K, y = Entropy)) + geom_path() +
  xlab('Number of topics') +
  scale_y_continuous(limits = c(0,1)) +
  theme_bw()

ggsave("study2_entropies.png", p, device = "png")
ggsave("study2_entropies.svg", p, device = "svg")

p <- ggplot(data.frame(K = seqk, ll = ll), aes(x = K, y = ll)) + geom_path() +
  geom_vline(xintercept = (which.max(ll)+1), linetype = 2) +
  xlab('Number of topics') +
  geom_smooth(method = "lm", formula = y~log(x), se = FALSE)+
  theme_bw()

ggsave("study2_ll.png", p, device = "png")
ggsave("study2_ll.svg", p, device = "svg")

p <- ggplot(data.frame(K = seqk, BIC = bics), aes(x = K, y = BIC)) + geom_path() +
  geom_vline(xintercept = (which.min(bics)+1), linetype = 2) +
  xlab('Number of topics') +
  geom_smooth(method = "lm", formula = y~x, se = FALSE)+
  theme_bw()

ggsave("study2_BIC.png", p, device = "png")
ggsave("study2_BIC.svg", p, device = "svg")
lda_dims_2 <- read_yaml("Study2_lda_dims.txt")
```
The corpus for this second analysis consisted of the abstracts of the selected articles.
To perform feature extraction,
we first applied the natural language processing technique "part-of-speech tagging" (POS-tagging), which identifies a word's grammatical function within the sentence context.
Because our analysis sought to identify phenomena,
we retained only nouns (to capture terms like "emotion") and adjectives (to capture the "mental" in "mental health").
Retaining nouns and adjectives generally helps derive more interpretable text mining models [@martinMoreEfficientTopic2015].
Finally, we used stemming to reduce the retained terms to their root form.
The resulting corpus consisted of 15587 terms in 4414 documents.
